---
title: "analysis"
output: html_document
---

# Final Project - Group 6

Based on: <https://github.com/21chanas3/dsci100-group73>

**Using clinical features to predict likelihood of a stroke in patients**

Original Authors: Aden Chan, Matthew Chan, Shadib Hossain, Brian Vu

Modified By: Aden Chan, Duncan Harrop, Navya Sehgal, Una Chou

# Abstract

This project investigates whether a patient's likelihood of stroke can be predicted using a set of clinical features from the Stroke Prediction Dataset. The dataset contains 5,110 observations and includes variables such as age, gender, hypertension, heart disease, residence type, average glucose level, BMI, smoking status, work type, marital status, and prior stroke occurrence. After cleaning and wrangling the data, we conducted exploratory data analysis to identify patterns between these predictors and stroke outcomes, then developed a k-nearest neighbors classification model using an 80/20 train-test split and 5-fold cross-validation for model tuning.

Our exploratory analysis suggested that age, hypertension, heart disease, BMI, and average glucose level were the most informative predictors, while variables such as id, work type, and marital status contributed little to model performance. The final model achieved high overall accuracy on the test set, but the dataset was heavily imbalanced and the model classified nearly all cases as "No stroke," failing to meaningfully identify positive stroke cases.

These findings suggest that future work should consider alternative modeling strategies, better methods for handling class imbalance, and additional predictors related to lifestyle, genetics, and medical history.

# Introduction

According to the National Cancer Institute, a stroke occurs when brain tissue is damaged by a loss of blood flow to certain parts of the brain (NIH). It is a significant global health issue, with an annual mortality rate of 5.5 million people worldwide, making it the second leading cause of death. Understanding the factors that contribute to stroke occurrence is crucial for prevention and effective treatment. Age is widely recognized as the strongest determinant of stroke risk, with the likelihood of experiencing a stroke doubling every decade after the age of 55. Additionally, hypertension has been identified as the leading risk factor of stroke in both developing and developed nations (Donker, 2018).

However, it is worth noting that there are other risk factors that exist, some of which are challenging to quantify accurately. These factors contribute to the complexity of predicting likelihood of a stroke. In light of this, our research question aims to address this challenge by **exploring the ability to accurately predict the likelihood of a stroke in a patient given a set of clinical features**.

To answer this research question, we will be utilizing the Stroke Prediction Dataset, available via Kaggle. The dataset comprises 5110 observations and contains 12 different variables which make up our clinical features for patients. These variables include gender, age, hypertension, heart disease, marriage status, type of work (self-employed/private/children/government/never worked), residence (rural or urban), average glucose levels, BMI, smoking status, and whether the patient has ever had a stroke.

By leveraging this dataset, we aim to identify patterns among these clinical features and the occurrence of stroke. This analysis will contribute to our understanding of the predictive power of these factors and potentially aid in the development of effective strategies for stroke prevention and management.

# Methods and Results

```{r setup, include=FALSE}
library(MASS)
library(tidyverse)
library(tidymodels)
library(repr)
library(gridExtra)
library(GGally)
library(broom)
library(ranger)
library(scales)
library(grid)
library(themis)
library(vip)
library(finetune)
library(xgboost)
library(glmnet)


set.seed(894235)

plot_theme <- theme_bw(base_size = 11) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    strip.text = element_text(size = 9)
  )

  panel_theme <- plot_theme +
  theme(plot.margin = margin(5.5, 15, 5.5, 5.5))
```

## Loading Data

```{bash}
mkdir -p data

curl -L -o data/stroke-prediction-dataset.zip\
  https://www.kaggle.com/api/v1/datasets/download/fedesoriano/stroke-prediction-dataset
  
unzip -o data/stroke-prediction-dataset.zip -d data/

rm data/stroke-prediction-dataset.zip
```

```{r}
stroke <- read_csv("data/healthcare-dataset-stroke-data.csv")
```

We will next clean/wrangle our data into a tidy format - this involves the following steps: renaming of columns, converting the string “Unknown” into NA values, converting categorical variables into factors, renaming these factors into human-readable names, converting columns that should be numerical into doubles. We will drop the rows with *NA* values from the dataset, so that we continue analysis only will fully-formed data.

```{r}
#Rename columns to all lowercase
stroke_colnames <- stroke |> colnames() |> make.names() |> tolower()
colnames(stroke) <- stroke_colnames

#Convert Unknown's to NA
stroke <- stroke |> mutate(smoking_status = na_if(smoking_status, "Unknown"))

#Convert categorical vars to factors
stroke <- stroke |> mutate(gender = as_factor(gender), 
                 work_type = as_factor(work_type), 
                 residence_type = as_factor(residence_type), 
                 smoking_status = as_factor(smoking_status),
                 hypertension = as_factor(hypertension),
                 ever_married = as_factor(ever_married),
                 heart_disease = as_factor(heart_disease),
                 stroke = as_factor(stroke))

#Rename factor levels to nicer names
stroke$hypertension <- recode_factor(stroke$hypertension, "0" = "No", "1" = "Yes")
stroke$heart_disease <- recode_factor(stroke$heart_disease, "0" = "No", "1" = "Yes")
stroke$stroke <- recode_factor(stroke$stroke, "0" = "No", "1" = "Yes")
stroke$work_type <- recode_factor(stroke$work_type, "Govt_job" = "Government", "Never_worked" = "Never Worked")
stroke$smoking_status <- recode_factor(stroke$smoking_status, "formerly smoked" = "Formerly", "never smoked" = "Never", "smokes" = "Smokes")

#For some reason bmi is of type char probably because of N/A, converting to double here
stroke <- stroke |> mutate(bmi = as.numeric(as.character(bmi)))

#Dropping NA
stroke <- stroke |> drop_na()

#Removing the 1 other observation
stroke <- stroke |>
  filter(gender != "Other")
```

To predict whether a patient will experience a stroke, we will develop a k-nearest neighbors classification model. To do so, we need to split our dataset using part of it to train the model and the other part of it to test the model. The training data will be used to both select the features used in the final model and to tune the number of neighbors used in classification, and the testing data will be used to evaluate the model’s accuracy on unseen data.

We will select an 80/10/10 split for training, validation, and test data, and will set `strata=stroke` to ensure that all sets have a proportionate amount of the outcome we are predicting.

```{r}
#Creating test/train split here
stroke_split <- initial_validation_split(stroke, prop = c(0.8, 0.1), strata = stroke)
stroke_training <- training(stroke_split)
stroke_validation <- validation(stroke_split)
stroke_testing <- testing(stroke_split)
```

Our exploratory data analysis will exclusively use our training data set and begins by obtaining summary statistics for our categorical variables and numeric variables. For our categorical variables, we obtain a count of each category - for example, `ever_married` has two values: `{Yes, No}` and our training set has 2084 yes and 656 no.

```{r}
#Getting summary stats for categorical vars here
stroke_married_stats <- stroke_training |> group_by(ever_married) |> summarise(value = n()) |> 
    rename("key" = "ever_married") |> mutate(column_name = "ever_married")
stroke_work_stats <- stroke_training |> group_by(work_type) |> summarise(value = n()) |> 
    rename("key" = "work_type") |> mutate(column_name = "work_type")
stroke_gender_stats <- stroke_training |> group_by(gender) |> summarise(value = n()) |> 
    rename("key" = "gender") |> mutate(column_name = "gender")
stroke_residence_type_stats <- stroke_training |> group_by(residence_type) |> summarise(value = n()) |> 
    rename("key" = "residence_type") |> mutate(column_name = "residence_type")
stroke_smoking_status_stats <- stroke_training |> group_by(smoking_status) |> summarise(value = n()) |> 
    rename("key" = "smoking_status") |> mutate(column_name = "smoking_status")
stroke_hypertension_stats <- stroke_training |> group_by(hypertension) |> summarise(value = n()) |> mutate(hypertension = as.character(hypertension)) |> 
    rename("key" = "hypertension") |> mutate(column_name = "hypertension") 
stroke_heart_disease_stats <- stroke_training |> group_by(heart_disease) |> summarise(value = n()) |> mutate(heart_disease = as.character(heart_disease)) |> 
    rename("key" = "heart_disease") |> mutate(column_name = "heart_disease")
stroke_stroke_stats <- stroke_training |> group_by(stroke) |> summarise(value = n()) |> mutate(stroke = as.character(stroke)) |> 
    rename("key" = "stroke") |> mutate(column_name = "stroke") 
```

For our numeric values, we obtain the mean.

```{r}
#Getting summary stats for continuous vars here
stroke_numeric_cols_summary <- stroke |> 
    select(avg_glucose_level, bmi, age) |> 
    map_dfc(mean, na.rm = TRUE)
```

Combine both the categorical and numeric statistics into a table.

```{r}
#Combining both summary stats
stroke_summary_stats <- stroke_numeric_cols_summary |> pivot_longer(age:avg_glucose_level, names_to = "column_name", values_to = "value")
stroke_summary_stats <- stroke_summary_stats |> mutate(key = "mean")
stroke_summary_stats <- bind_rows(stroke_summary_stats, 
                                  stroke_gender_stats, 
                                  stroke_residence_type_stats,
                                  stroke_smoking_status_stats,
                                  stroke_stroke_stats,
                                  stroke_hypertension_stats,
                                  stroke_heart_disease_stats,
                                  stroke_work_stats,
                                  stroke_married_stats)
stroke_summary_stats <- stroke_summary_stats[, c(1,3,2)]
stroke_summary_stats |> as_tibble() |> print(n = 25)
```

## Visualizations and Justification for feature selection

Next, we need to determine which of the original datasets’ features will be useful in determining whether or not a patient will experience a stroke. This is called exploratory data analysis and we begin by plotting our features to visually determine if there is a correlation between the feature and the probability of a stroke.

The first feature of the original dataset that will be removed from analysis is `id` - this simply corresponds to a patient's ID and has no impact on whether or not a patient will have a stroke.

```{r}
stroke_training <- stroke_training |> select(-id)
```

### Preliminary analysis

To begin our analysis, we employed the ggpairs tool, enabling us to gain a holistic view of the dataset and its clinical variables. By generating a matrix where each variable was plotted against one another with whether a patient had a stroke before, we could visually identify variables that had correlations with one another or with stroke. Consequently, this provided us with valuable insights into the variables that needed further investigation below and those that could be excluded due to their lack of significance in our analysis.

```{r}
options(repr.plot.width = 12, repr.plot.height = 10)

stroke_pairs <- stroke |>
  select(
    stroke,
    age,
    avg_glucose_level,
    bmi,
    hypertension,
    heart_disease
  )

plot_pairs <- ggpairs(
  stroke_pairs,
  columns = 2:6,
  mapping = aes(color = stroke, alpha = 0.5),
  upper = list(
    continuous = wrap("points", size = 0.4)
  ),
  lower = list(
    continuous = wrap("smooth", se = FALSE, alpha = 0.3)
  ),
  diag = list(
    continuous = wrap("densityDiag", alpha = 0.5)
  )
) +
  labs(
    title = "Figure 1. Pairwise relationships among selected stroke predictors",
    color = "Stroke"
  ) +
  theme_bw(base_size = 10) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    legend.position = "bottom",
    axis.text.x = element_text(size = 8),
    axis.text.y = element_text(size = 8),
    strip.text = element_text(size = 9)
  )

suppressMessages(suppressWarnings(print(plot_pairs)))
```

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)

stroke_gender <- stroke |>
  count(gender, stroke) |>
  group_by(gender) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = gender, y = percentage, fill = stroke)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Gender",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 2. Percentage of stroke cases by gender"
  ) +
  plot_theme

stroke_gender
```

There was no difference between males and females in terms of the number of stroke cases.

```{r}
options(repr.plot.width = 10, repr.plot.height = 6)

stroke_age <- stroke |>
  ggplot(aes(x = age, fill = stroke)) +
  geom_histogram(
    position = "fill",
    binwidth = 5,
    color = "white"
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Age",
    y = "Proportion",
    fill = "Stroke",
    title = "Figure 3. Proportion of stroke cases by age group"
  ) +
  plot_theme

stroke_age
```

With regards to age, the proportion of stroke cases is very low among younger patients and increases noticeably at older ages. In particular, the proportion begins to rise after around age 50 and becomes much more pronounced among patients over 60. This suggests that age is one of the strongest predictors of stroke in the dataset.

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)

stroke_hypertension <- stroke |>
  count(hypertension, stroke) |>
  group_by(hypertension) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = hypertension, y = percentage, fill = stroke)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Hypertension",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 4. Percentage of stroke cases by hypertension"
  ) +
  plot_theme

stroke_hypertension
```

Those who have suffered hypertension are roughly twice as likely to have a stroke, which is around 12.5% of the cases.

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)

stroke_heart_disease <- stroke |>
  count(heart_disease, stroke) |>
  group_by(heart_disease) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = heart_disease, y = percentage, fill = stroke)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Heart disease",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 5. Percentage of stroke cases by heart disease"
  ) +
  plot_theme

stroke_heart_disease
```

Those who have suffered heart disease are roughly twice as likely to have a stroke, which is around 20-25% of the cases.

```{r}
options(repr.plot.width = 8, repr.plot.height = 5)

stroke_residence <- stroke |>
  count(residence_type, stroke) |>
  group_by(residence_type) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = residence_type, y = percentage, fill = stroke)) +
  geom_col(width = 0.75) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Type of residence",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 6. Percentage of stroke cases by type of residence"
  ) +
  plot_theme

stroke_residence
```

There is no difference between those who live in urban areas and those who live in more rural areas and their likelihood of getting a stroke.

```{r}
options(repr.plot.width = 8, repr.plot.height = 6)

stroke_avg_glucose_level <- stroke |>
  ggplot(aes(x = stroke, y = avg_glucose_level, fill = stroke)) +
  geom_boxplot(alpha = 0.8, width = 0.6) +
  labs(
    x = "Stroke",
    y = "Average glucose level",
    fill = "Stroke",
    title = "Figure 7. Average glucose level by stroke status"
  ) +
  plot_theme +
  theme(legend.position = "none")

stroke_avg_glucose_level
```

There is an association between stroke cases and average glucose level. Patients who had experienced a stroke tend to have higher glucose levels on average than those who had not. While there is overlap between the two groups, the stroke group shows a higher center and a wider spread.


```{r}
options(repr.plot.width = 8, repr.plot.height = 6)

stroke_bmi <- stroke |>
  ggplot(aes(x = stroke, y = bmi, fill = stroke)) +
  geom_boxplot(alpha = 0.8, width = 0.6, na.rm = TRUE) +
  labs(
    x = "Stroke",
    y = "BMI",
    fill = "Stroke",
    title = "Figure 8. BMI by stroke status"
  ) +
  plot_theme +
  theme(legend.position = "none")

stroke_bmi
```

BMI appears to be somewhat associated with stroke status, although the separation is weaker than for age. Patients with a prior stroke tend to have slightly higher BMI values on average, but there is substantial overlap between the two groups.

```{r}
options(repr.plot.width = 9, repr.plot.height = 5)

stroke_smoking_status <- stroke |>
  count(smoking_status, stroke) |>
  group_by(smoking_status) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = smoking_status, y = percentage, fill = stroke)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Smoking status",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 9. Percentage of stroke cases by smoking status"
  ) +
  plot_theme +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

stroke_smoking_status
```

The percentage of cases are quite similar between the three groups. However, those who have formerly smoked have a slightly higher risk of stroke than those who smoke. This may be due to current smokers being younger or having less overall damage than those who are former smokers. Potentially, those who are former smokers may have been heavy smokers which could explain why they quit smoking compared to those who currently smoke. Interestingly, those who have never smoked have the same likelihood of suffering a stroke as those who currently smoke. This means that other lifestyle factors likely play a significant role in the risk of getting a stroke.

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)

stroke_worktype_plot <- stroke |>
  count(work_type, stroke) |>
  group_by(work_type) |>
  mutate(percentage = n / sum(n)) |>
  ggplot(aes(x = work_type, y = percentage, fill = stroke)) +
  geom_col() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Type of work",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 10. Percentage of stroke cases by work type"
  ) +
  plot_theme +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

stroke_worktype_plot
```

Those who are self-employed are roughly twice as likely to suffer a stroke than those who work in government or in the private sector, at around 20% of cases. Potentially, this is due to self employed participants being in more incidents where their blood pressure is high because of the uncertainties that are associated with being in that work type.

### Further Analysis

To further explore the relationships between stroke and various clinical variables, we conducted an analysis focusing on two key variables in relations to strokes. By examining their associations with stroke, we aimed to uncover potential interrelations and gain deeper insights into their significance.

```{r}
options(repr.plot.height = 5, repr.plot.width = 14)

stroke_age_gender <- stroke |>
  ggplot(aes(x = age, fill = stroke)) +
  geom_histogram(
    position = "fill",
    binwidth = 5,
    color = "white"
  ) +
  facet_wrap(~ gender) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = "Age",
    y = "Proportion",
    fill = "Stroke",
    title = "Figure 11. Proportion of stroke cases by age group and gender"
  ) +
  plot_theme

stroke_age_gender
```

While in our initial analysis, Gender was found to not be a good predictor of strokes, we decided to see if there was a correlation between Age, Gender, and Strokes. It appears that the risk of stroke is slightly greater in older men than older women; however, age seems to be the more dominant feature. We decided to still consider gender in our model as we were interested in this difference as there is a chance some participants in the female population were pregnant, making blood more likely clot which could cause a stroke.

```{r}
options(repr.plot.width = 12, repr.plot.height = 6)

stroke_age_smoking <- stroke %>%
  ggplot(aes(x = age, fill = stroke)) +
  geom_histogram(
    position = "fill",
    binwidth = 5,
    color = "white"
  ) +
  facet_wrap(~ smoking_status) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Age",
    y = "Proportion",
    fill = "Stroke",
    title = "Figure 12. Proportion of stroke cases by age group and smoking status"
  ) +
  plot_theme

stroke_age_smoking
```

All smoking groups had roughly equal numbers of stroke cases with the common ground being that with age, there was an average increase in percentage of strokes. The main difference is that the peak percentage was highest in the group that currently “smokes” at around 50%. The “never” smoking and “formerly” smoking groups peaked at percentages between 40% to 50%. Interestingly, former smokers had higher percentages than those who never smoked. Both of those groups had lower percentages in comparison to current smokers, especially as age increases. This suggests a relationship that has been observed in the past, where stopping smoking for 5 years reduced the risk of stroke comparable to that of a nonsmoker in most cases (Shah & Cole, 2010).

```{r}
options(repr.plot.width = 8, repr.plot.height = 6)

heartdisease_hypertension <- stroke %>%
  count(hypertension, heart_disease, stroke) %>%
  group_by(hypertension, heart_disease) %>%
  mutate(percentage = n / sum(n)) %>%
  ungroup()

heartdisease_hypertension_plot <- ggplot(
  heartdisease_hypertension,
  aes(x = hypertension, y = percentage, fill = stroke)
) +
  geom_col(position = "stack") +
  facet_wrap(~ heart_disease) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Hypertension",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 13. Percentage of stroke cases by hypertension and heart disease"
  ) +
  plot_theme

heartdisease_hypertension_plot
```

In general, individuals without hypertension and heart disease have the lowest likelihood of having had a stroke, with a prevalence of less than 5%. However, those with both hypertension and heart disease exhibit the highest percentage, around 18%, of stroke cases. Participants with only heart disease have a lower likelihood, less than 18%, while those with only hypertension have an approximate 13% likelihood of having had a stroke.

```{r}
options(repr.plot.width = 10, repr.plot.height = 5)

heartdisease_gender <- stroke %>%
  count(gender, heart_disease, stroke) %>%
  group_by(gender, heart_disease) %>%
  mutate(percentage = n / sum(n))

stroke_gender_plot <- heartdisease_gender %>%
  ggplot(aes(x = heart_disease, y = percentage, fill = stroke)) +
  geom_col() +
  facet_wrap(~ gender) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    x = "Heart disease",
    y = "Percentage",
    fill = "Stroke",
    title = "Figure 14. Percentage of stroke cases by gender and heart disease"
  ) +
  plot_theme

stroke_gender_plot
```

While there isn't a strong association between heart disease and gender, both male and female participants with heart disease have a notably higher likelihood of having had a stroke compared to those without a heart condition.

```{r}
options(repr.plot.width = 8, repr.plot.height = 6)

age_bmi <- stroke %>%
  ggplot(aes(x = age, y = bmi, colour = stroke)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "Age",
    y = "BMI",
    color = "Stroke",
    title = "Figure 15. Stroke cases by age and BMI"
  ) +
  plot_theme +
  scale_color_manual(
    values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0))
  )

age_bmi
```

Age seems to be the main predictor of stroke when BMI is roughly under 25. However, this is not for a majority of the cases. Most cases where participants have suffered a stroke, they tend to be above 50 years old and have a BMI of greater than 25.

```{r}
glucose_bmi <- stroke %>%
  ggplot(aes(x = bmi, y = avg_glucose_level, colour = stroke)) +
  geom_point(alpha = 0.5) +
  labs(
    x = "BMI",
    y = "Average glucose level",
    color = "Stroke",
    title = "Figure 16. Stroke cases by average glucose level and BMI"
  ) +
  plot_theme +
  scale_color_manual(values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0)))

glucose_bmi
```

Our analysis suggests that participants with a BMI of 25 or higher make up a majority of the instances of strokes regardless of average glucose levels. The majority of participants who did not suffer a stroke had average glucose levels of under 150. This tells us that BMI is a stronger predictor of whether someone had a stroke, but maintaining glucose levels at under 150 and BMI under 25 may reduce the risks of stroke.

```{r}
options(repr.plot.width = 15, repr.plot.height = 6)

glucose_hypertension <- stroke %>%
  ggplot(aes(x = avg_glucose_level, y = hypertension, colour = stroke)) +
  geom_jitter(height = 0.15, width = 0, alpha = 0.5) +
  labs(
    title = "A. Hypertension",
    x = "Average glucose level",
    y = "Hypertension",
    color = "Stroke"
  ) +
  plot_theme +
  theme(legend.position = "none") +
  scale_color_manual(
    values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0))
  )

glucose_heartdisease <- stroke %>%
  ggplot(aes(x = avg_glucose_level, y = heart_disease, colour = stroke)) +
  geom_jitter(height = 0.15, width = 0, alpha = 0.5) +
  labs(
    title = "B. Heart disease",
    x = "Average glucose level",
    y = "Heart disease",
    color = "Stroke"
  ) +
  plot_theme +
  scale_color_manual(
    values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0))
  )

grid.arrange(
  glucose_hypertension,
  glucose_heartdisease,
  ncol = 2,
  top = textGrob(
    "Figure 17. Stroke cases by glucose level with hypertension and heart disease",
    gp = gpar(fontsize = 12)
  )
)
```

**Stroke Cases by Average Glucose Levels and Hypertension**

Most observations where the participant did not suffer a stroke are associated with not having hypertension and with average glucose levels below approximately 125. Stroke cases appear somewhat more frequently among participants with higher glucose levels, but there remains substantial overlap between stroke and non-stroke observations. Participants with hypertension show a similar proportion of stroke cases across a range of glucose levels, suggesting that once hypertension is present, glucose level alone may not strongly differentiate stroke risk.

**Stroke Cases by Average Glucose Levels and Heart Disease**

A similar pattern appears when examining heart disease. Most non-stroke observations occur among participants without heart disease and with lower average glucose levels. Stroke cases appear somewhat more frequently at higher glucose levels, although the separation between stroke and non-stroke observations remains limited. Participants with heart disease exhibit a slightly higher proportion of stroke cases overall, but glucose level alone does not clearly distinguish between stroke and non-stroke outcomes.

```{r}
options(repr.plot.width = 15, repr.plot.height = 6)

residence_heartdisease <- stroke %>%
  ggplot(aes(x = residence_type, y = heart_disease, colour = stroke)) +
  geom_jitter(width = 0.15, height = 0.15, alpha = 0.5) +
  labs(
    title = "A. Heart disease",
    x = "Type of residence",
    y = "Heart disease",
    color = "Stroke"
  ) +
  plot_theme +
  theme(legend.position = "none") +
  scale_color_manual(
    values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0))
  )

residence_hypertension <- stroke %>%
  ggplot(aes(x = residence_type, y = hypertension, colour = stroke)) +
  geom_jitter(width = 0.15, height = 0.15, alpha = 0.5) +
  labs(
    title = "B. Hypertension",
    x = "Type of residence",
    y = "Hypertension",
    color = "Stroke"
  ) +
  plot_theme +
  scale_color_manual(
    values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0))
  )

grid.arrange(
  residence_heartdisease,
  residence_hypertension,
  ncol = 2,
  top = textGrob(
    "Figure 18. Stroke cases by residence type with heart disease and hypertension",
    gp = gpar(fontsize = 12)
  )
)
```

```{r}
options(repr.plot.width = 15, repr.plot.height = 6)

heartdisease_residence <- stroke %>%
  count(residence_type, heart_disease, stroke) %>%
  group_by(residence_type, heart_disease) %>%
  mutate(percentage = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = residence_type, y = percentage, fill = stroke)) +
  geom_col() +
  facet_wrap(~ heart_disease) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "A. Heart disease",
    x = "Type of residence",
    y = "Percentage",
    fill = "Stroke"
  ) +
  plot_theme +
  theme(legend.position = "none")

hypertension_residence <- stroke %>%
  count(residence_type, hypertension, stroke) %>%
  group_by(residence_type, hypertension) %>%
  mutate(percentage = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = residence_type, y = percentage, fill = stroke)) +
  geom_col() +
  facet_wrap(~ hypertension) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "B. Hypertension",
    x = "Type of residence",
    y = "Percentage",
    fill = "Stroke"
  ) +
  plot_theme

grid.arrange(
  heartdisease_residence,
  hypertension_residence,
  ncol = 2,
  top = textGrob(
    "Figure 19. Stroke percentages by residence, heart disease, and hypertension",
    gp = gpar(fontsize = 12)
  )
)
```

[**Stroke Cases by Type of Residence and Heart Disease**]{.underline}

The instances of stroke are roughly equal for those who have not experienced heart disease, regardless of their type of residence. Cases of strokes are also roughly equal between those who have experienced heart disease, regardless of their residence type. Most cases of no stroke are prevalent in those who did not experience heart disease and overall, there is a higher occurrence of strokes for those who have experienced heart disease.

[**Stroke Cases by Residence Type and Hypertension**]{.underline}

Regardless of one’s residence type, cases of strokes are roughly equal when one experiences hypertension. The prevalence of strokes in those who did not have hypertension is also equal in both urban and rural areas, and most cases of no stroke are present in those who do not have hypertension.

This suggests that hypertension and heart disease are likely a stronger predictor of stroke than the type of residence one resides in.

## Building The Model

### Finalized Features

As a result of our data analysis, we will be moving forward with the following features as predictors of stroke.

| Predictor             | Definition                                     |
|-----------------------|------------------------------------------------|
| Gender                | Whether the person is male or female           |
| Age                   | Age of patient                                 |
| Hypertension          | Diagnosed with hypertension or not             |
| Heart disease         | Diagnosed with heart disease or not            |
| Residence             | Living in a rural or urban area                |
| Average glucose level | Average glucose level in blood (mg/dL)         |
| BMI                   | Body mass index                                |
| Smoking status        | Current smoker, former smoker, or never smoked |
| Stroke                | Whether the person has had a stroke or not     |

And these are the features we will not be using in our analysis:

| Unused Predictor | Definition | Exclusion Justification |
|---------------------|--------------------|-------------------------------|
| id | ID of patient | Does not contribute to stroke prediction |
| ever_married | Ever married or not | This is too broad of a predictor\* |
| work_type | Either "Private", "Self-Employed", "Government job", "Never worked", "Children" | Options are too broad\*\* |

\*There were no research findings indicating that marital status alone is predictive of stroke. Additionally, the variable does not account for marital satisfaction, which at least for men has been shown to increase long-term risk of stroke and all-cause mortality when satisfaction is low (Lev-Ari, Gepner, & Goldbourt, 2021).

\*\*We believe the `work_type` categories are not sufficiently detailed to accurately predict stroke risk. Important socioeconomic factors such as income level or occupational stress are not captured. Research has shown that higher socioeconomic status is positively related to stroke risk (Wang et al., 2019). Additionally, studies suggest that white-collar workers may have a higher risk of hypertension, a major risk factor for stroke, compared to blue-collar workers (Liu et al., 2017).

### kNN Classification

Now that we have finalized the features that will be used in the model, we define the final `recipe` that will be used. The predictors are scaled as described above.

```{r}
stroke_recipe <- recipe(stroke ~ gender+age+hypertension+heart_disease+residence_type+avg_glucose_level+bmi+smoking_status, data = stroke_training) |> 
            step_YeoJohnson(all_numeric_predictors()) |>
            step_scale(all_numeric_predictors()) |> 
            step_center(all_numeric_predictors()) |>
            step_dummy(all_nominal_predictors()) |>
            step_smote(stroke) 
```

#### Hyperparameter Tuning

We will tune the neighbors parameter via 5-fold cross-validation with a coarse sweep from 1 to 50 neighbors, incrementing by 5 neighbors, followed by a finer sweep. We will use Youden's J statistic to evaluate model performance as accuracy may improperly "mask" a poorly predictive model due to the class imbalance present. Youden's J statistic accounts both for sensitivity and specificity.

```{r}
stroke_knn_classifier <- nearest_neighbor(weight_func = "rectangular", 
                                          neighbors = tune()) |> 
  set_engine("kknn") |> 
  set_mode("classification")

stroke_knn_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)
coarse_knn_k_vals <- tibble(neighbors = seq(from = 1, to = 200, by = 10))

coarse_knn_cv_sweep_results <- workflow() |>
  add_recipe(stroke_recipe) |>
  add_model(stroke_knn_classifier) |>
  tune_grid(resamples = stroke_knn_vfold, 
            grid      = coarse_knn_k_vals,
            metrics   = metric_set(j_index, sensitivity, accuracy)) |>
  collect_metrics()
coarse_knn_cv_sweep_results |> 
      filter(.metric == "j_index") |>
      arrange(desc(mean))
```

```{r}
coarse_n_neighbors_plot <- coarse_knn_cv_sweep_results |>
  filter(.metric == "j_index") |>
  ggplot(aes(x=neighbors, y=mean)) +
  geom_line() +
  labs(title = "Figure 20. J-Index vs # of Neighbours for kNN Model",
       x     = "# of Neighbours",
       y     = "Mean J-Index")

coarse_n_neighbors_plot
```

From Figure 20, we can see that model performance increases steadily as the number of neighbors increases, reaching a plateau around 150. We will conduct a finer search around the 150-180 neighbour range to land on a final value.

```{r}
fine_knn_k_vals <- tibble(neighbors = seq(from = 150, to = 180, by = 1))

fine_knn_cv_sweep_results <- workflow() |>
  add_recipe(stroke_recipe) |>
  add_model(stroke_knn_classifier) |>
  tune_grid(resamples = stroke_knn_vfold, 
            grid      = fine_knn_k_vals,
            metrics   = metric_set(j_index, sensitivity, accuracy)) |>
  collect_metrics()
fine_knn_cv_sweep_results |> 
      filter(.metric == "j_index") |>
      arrange(desc(mean))
```

It was found that `k=159` results in the highest score on our training dataset. We will use this value to develop our final kNN model with `neighbors=159`.

#### Evaluation on Validation Set

```{r}
knn_model_final <- nearest_neighbor(weight_func = "rectangular", neighbors = 159) |>   
  set_engine("kknn") |>
  set_mode("classification")
```

We train our final model and fit it against our training dataset.

```{r}
knn_stroke_fit <- workflow() |> 
  add_model(knn_model_final) |> 
  add_recipe(stroke_recipe) |> 
  fit(data = stroke_training)
```

We used the model trained with the optimal number of neighbors, and then use it to predict new values using our validation set and calculate the J-index obtained

```{r}
knn_stroke_test_predictions <- predict(knn_stroke_fit, stroke_validation) |> 
  bind_cols(stroke_validation)
```

```{r}
knn_model_metrics <- knn_stroke_test_predictions |> 
  metric_set(accuracy, kap, j_index)(truth = stroke, estimate = .pred_class)
knn_model_metrics
```

```{r}
knn_confusion <- knn_stroke_test_predictions |>
             conf_mat(truth = stroke, estimate = .pred_class)
knn_confusion
```

We get a final score of 0.4001, which is not too good as the metric is on a scale from [0,1]. From the confusion matrix, we can see that the model is predicting a lot of false positives.

### Logistic Regression

Logistic regression is another model type that can be used for classification tasks. It models the log-odds of the outcome as a linear combination of predictors. We will additionally use backward stepwise selection to prune non-informative terms, and tune an elastic-net penalty to regularize the retained coefficients.

#### Backward Selection

We first define a recipe without SMOTE to maintain the class distribution in the original dataset during backward selection:

```{r}
logr_recipe_for_selection <- recipe(
    stroke ~ gender + age + hypertension + heart_disease +
             residence_type + avg_glucose_level + bmi + smoking_status,
    data = stroke_training
  ) |>
  step_YeoJohnson(all_numeric_predictors()) |>
  step_scale(all_numeric_predictors()) |>
  step_center(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors())

baked_train <- prep(logr_recipe_for_selection) |> bake(new_data = NULL)

full_glm <- glm(stroke ~ ., data = baked_train, family = binomial)
backward_glm <- MASS::stepAIC(full_glm, direction = "backward", trace = FALSE)
summary(backward_glm)
```

```{r}
survived_terms <- attr(terms(backward_glm), "term.labels")
cat(survived_terms, sep = "\n")
```

The backward selection procedure removes predictors that do not improve the AIC, and resulted in only the terms left above. We will use these selected terms in the final model.

#### Hyperparameter Tuning

We add SMOTE back in for the tuning phase. The elastic-net penalty and regularisation strength are tuned at the same time over a 10 by 10 grid using 5-fold cross-validation evaluated by J-index as in the kNN model:

```{r}
logr_recipe_final <- recipe(
    stroke ~ age + avg_glucose_level + hypertension + heart_disease +
             smoking_status,
    data = stroke_training
  ) |>
  step_YeoJohnson(all_numeric_predictors()) |>
  step_scale(all_numeric_predictors()) |>
  step_center(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_smote(stroke)

logr_spec <- logistic_reg(penalty = tune(), mixture = tune()) |>
  set_engine("glmnet") |>
  set_mode("classification")

logr_grid <- grid_regular(
  penalty(range = c(-5, 0)),
  mixture(range = c(0, 1)),
  levels = 10
)

logr_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)

logr_cv_results <- workflow() |>
  add_recipe(logr_recipe_final) |>
  add_model(logr_spec) |>
  tune_grid(
    resamples = logr_vfold,
    grid      = logr_grid,
    metrics   = metric_set(j_index, sensitivity, accuracy)
  ) |>
  collect_metrics()

best_logr_params <- logr_cv_results |>
  filter(.metric == "j_index") |>
  arrange(desc(mean)) |>
  slice(1)
best_logr_params
```

The best performing model had `penalty = 0.2782559` and `mixture = 0`, which corresponds to a pure ridge regression model. We will use these hyperparameters to create the final model to be tested on the validation set.

#### Evaluation on Validation Set

```{r}
logr_final_spec <- logistic_reg(
    penalty = best_logr_params$penalty,
    mixture = best_logr_params$mixture
  ) |>
  set_engine("glmnet") |>
  set_mode("classification")

logr_fit <- workflow() |>
  add_recipe(logr_recipe_final) |>
  add_model(logr_final_spec) |>
  fit(data = stroke_training)
```

```{r}
logr_val_predictions <- predict(logr_fit, stroke_validation) |>
  bind_cols(stroke_validation)

logr_val_metrics <- logr_val_predictions |>
  metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate = .pred_class)
logr_val_metrics
```

```{r}
logr_confusion <- logr_val_predictions |>
  conf_mat(truth = stroke, estimate = .pred_class)
logr_confusion
```

The logistic model resulted in a substantial improvement over the kNN model, achieving a J-index of 0.6091. The number of false positives is also reduced compared to the kNN model.

### XGBoost

XGBoost is a gradient-boosted tree ensemble that naturally captures non-linear relationships and feature interactions. We will first fit a preliminary model to rank feature importance, using the information to prune uninformative predictors, then tune the model's hyperparameters using Bayesian optimisation.

#### Feature Importance

```{r}
xgb_recipe_full <- recipe(
    stroke ~ gender + age + hypertension + heart_disease +
             residence_type + avg_glucose_level + bmi + smoking_status,
    data = stroke_training
  ) |>
  step_YeoJohnson(all_numeric_predictors()) |>
  step_scale(all_numeric_predictors()) |>
  step_center(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_smote(stroke)

xgb_spec_initial <- boost_tree(trees = 200) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_fit_initial <- workflow() |>
  add_recipe(xgb_recipe_full) |>
  add_model(xgb_spec_initial) |>
  fit(data = stroke_training)
```

```{r}
vip_plot <- xgb_fit_initial |>
  extract_fit_parsnip() |>
  vip(num_features = 15) +
  labs(title = "Figure 21. XGBoost Feature Importance") +
  plot_theme
vip_plot
```

Features with importance that are near zero mainly contribute noise without improving prediction. However, due to the dummy step on catagorical predictors, we can see their their importance is split across its levels and thus the importance of the predictor must be interpreted as the sum of its importances across its derived dummy variables.

Based on the importance ranking above, `age`, `avg_glucose_level`, `bmi`, `hypertension`, `heart_disease`, and `smoking_status` are informative while we can and drop categorical predictors that were consistently near the bottom such as `gender` .

#### Bayesian Hyperparameter Tuning

We tune six hyperparameters simultaneously using Bayesian optimisation (`tune_bayes`), which is more efficient than an exhaustive grid search for high-dimensional parameter spaces. We run 10 initial random evaluations followed by up to 30 Bayesian iterations, stopping early if 10 consecutive iterations fail to improve J-index:

```{r}
xgb_recipe_selected <- recipe(
    stroke ~ age + avg_glucose_level + bmi +
             hypertension + heart_disease + smoking_status + residence_type,
    data = stroke_training
  ) |>
  step_YeoJohnson(all_numeric_predictors()) |>
  step_scale(all_numeric_predictors()) |>
  step_center(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |>
  step_smote(stroke)

xgb_spec_tune <- boost_tree(
  trees          = tune(),
  tree_depth     = tune(),
  learn_rate     = tune(),
  min_n          = tune(),
  loss_reduction = tune(),
  sample_size    = tune()
) |>
  set_engine("xgboost") |>
  set_mode("classification")

xgb_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)

xgb_params <- workflow() |>
  add_recipe(xgb_recipe_selected) |>
  add_model(xgb_spec_tune) |>
  extract_parameter_set_dials() |>
  update(
    learn_rate = learn_rate(range = c(-3, -1)),
    trees      = trees(range = c(100, 1000))
  )

xgb_bayes_results <- workflow() |>
  add_recipe(xgb_recipe_selected) |>
  add_model(xgb_spec_tune) |>
  tune_bayes(
    resamples  = xgb_vfold,
    param_info = xgb_params,
    iter       = 30,
    initial    = 10,
    metrics    = metric_set(j_index, sensitivity, accuracy),
    control    = control_bayes(no_improve = 10, verbose = FALSE)
  )

best_xgb_params <- xgb_bayes_results |>
  collect_metrics() |>
  filter(.metric == "j_index") |>
  arrange(desc(mean)) |>
  slice(1)
best_xgb_params
```

The best performing model was obtained with the hyperparameters listed above and will be used to train the final model.

#### Evaluation on Validation Set

```{r}
xgb_final_workflow <- workflow() |>
  add_recipe(xgb_recipe_selected) |>
  add_model(xgb_spec_tune) |>
  finalize_workflow(best_xgb_params)

xgb_fit <- xgb_final_workflow |>
  fit(data = stroke_training)
```

```{r}
xgb_val_predictions <- predict(xgb_fit, stroke_validation) |>
  bind_cols(stroke_validation)

xgb_val_metrics <- xgb_val_predictions |>
  metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate = .pred_class)
xgb_val_metrics
```

```{r}
xgb_confusion <- xgb_val_predictions |>
  conf_mat(truth = stroke, estimate = .pred_class)
xgb_confusion
```

## Model Selection

We compare the three models on the validation set using J-index, sensitivity, and accuracy.

```{r}
all_val_metrics <- bind_rows(
  knn_stroke_test_predictions |>
    metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate = .pred_class) |>
    mutate(model = "kNN"),
  logr_val_predictions |>
    metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate = .pred_class) |>
    mutate(model = "Logistic Regression"),
  xgb_val_predictions |>
    metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate = .pred_class) |>
    mutate(model = "XGBoost")
)
```

```{r}
metrics_plot <- all_val_metrics |>
  mutate(
    .metric = recode(.metric,
      "j_index"     = "J-Index",
      "sensitivity" = "Sensitivity",
      "accuracy"    = "Accuracy"
    ),
    model = factor(model, levels = c("kNN", "Logistic Regression", "XGBoost"))
  ) |>
  ggplot(aes(x = .metric, y = .estimate, fill = model)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Figure 22. Validation Set Metrics by Model",
    x     = "Metric",
    y     = "Estimate",
    fill  = "Model"
  ) +
  plot_theme
metrics_plot
```

```{r}
knn_cm_plot  <- autoplot(knn_confusion,  type = "heatmap") +
  labs(title = "kNN") +
  plot_theme +
  theme(legend.position = "none")

logr_cm_plot <- autoplot(logr_confusion, type = "heatmap") +
  labs(title = "Logistic Regression") +
  plot_theme +
  theme(legend.position = "none")

xgb_cm_plot  <- autoplot(xgb_confusion,  type = "heatmap") +
  labs(title = "XGBoost") +
  plot_theme +
  theme(legend.position = "none")

confusion_grid <- grid.arrange(
  knn_cm_plot, logr_cm_plot, xgb_cm_plot,
  ncol   = 3,
  top    = grid::textGrob(
    "Figure 23. Confusion Matrices on Validation Set",
    gp = grid::gpar(fontsize = 11)
  )
)
```

Looking at the three models, logistic regression performed the best in terms of J-Index, at 0.61534. Inspecting the confusion matrices, we can see that it performed better than the kNN model at avoiding false positive and false negatives, but only outperformed the XGBoost model in avoiding false negatives. As the J-index metric weights both sensitivity and specificity equally, we will select the logistic regression model as our final candidate.

## Final Evaluation

```{r}
logr_test_predictions <- predict(logr_fit, stroke_testing) |>
  bind_cols(stroke_testing)

logr_test_metrics <- logr_test_predictions |>
  metric_set(j_index, sensitivity, specificity, accuracy)(truth = stroke, estimate = .pred_class)
logr_test_metrics
```

```{r}
logr_test_confusion <- logr_test_predictions |>
  conf_mat(truth = stroke, estimate = .pred_class)
logr_test_confusion
```

The performance of the logistic regression model on the testing set yielded a J-Index of 0.5556. This represents a drop as compared to the validation set, which mday indicate the model is having difficulty generalizing to unseen data.

# Discussion

The research project aimed to accurately predict the occurrence of stroke in patients based on a set of clinical features. The researchers utilized the Stroke Prediction Dataset, which included information on 12 variables, such as age, gender, hypertension, heart disease, average glucose levels, BMI, smoking status, and more. By analyzing this dataset, the researchers identified patterns between these clinical features and the occurrence of stroke.

Through backwards selection, age (p \< 2e-16), average glucose level (0.0047), and the presence of pre-existing heart disease emerged as the strongest predictors of stroke for our logistic regression model. Age was significant in determining stroke risk, with the likelihood of experiencing a stroke doubling every decade after the age of 55. Heart disease and average glucose levels were also identified as significant (p = 4.7e-3 and p = 2e-2 respectively) risk factors for stroke. Other factors, such as heart disease, BMI, smoking status, and type of work, showed varying degrees of association with stroke occurrence.

Visualizations and exploratory data analysis played a crucial role in identifying correlations between the clinical variables and stroke. For instance, it was observed that individuals above the age of 60 were approximately twice as likely to suffer a stroke compared to those in their mid to late 40s. Similarly, participants with hypertension or heart disease were at an increased risk of stroke. BMI and average glucose levels were found to have a positive association with stroke cases.

However, the research did not yield the expected results. The occurance of false positives are quite high, and the model also predicted some false negatives. In the context of stroke risk, false positives are less serious than false negatives, while the former would still result in unncessary usage of medical resources and costs, the latter could potentially have serious or lethal consequences if the model were to be deployed. These findings suggest that to develop a more accurate model, additional factors beyond those present in our dataset, such as genetic factors/family history and lifestyle choices, need to be considered. Additionally, metrics that penalize false positives more severely such as F2 could be used.

## Potential Limitations

Our current classification model, based on a regularized logistic regression model, has certain limitations that should be considered. Instead of predicting stroke occurrence, it focuses on determining whether a person has had a stroke based on the predictors utilized in our analysis. One limitation is that the dataset used represents a snapshot of stroke cases at a specific point in time. This can potentially limit its generalizability to future cases. Changes in stroke occurrence after the data collection period could affect the predictive power of the model. Additionally, our model does not account for potential risk factors that were not included in the dataset but may contribute to stroke occurrence. Their absence in the analysis may affect the accuracy of our predictions. To address these limitations, it may be beneficial to incorporate data from before the stroke event and establish a timeline to track the changes in clinical features over time. This approach could enhance our understanding of stroke occurrence and improve the accuracy of future classification models.

## Impacts

### Early Intervention

By being able to predict stroke risks, healthcare providers can better identify individuals who are at a higher risk of stroke before the deadly event happens. Early identification has the potential to significantly reduce the occurrence of strokes and further complications by allowing individuals to develop personalized treatment plans with medical professionals to improve their quality of life.

### Education and Awareness

Our findings can be used to educate future generations of the risk factors that are associated with having a stroke. This can encourage them to live a healthier and more positive lifestyle. As well, this may offer those who have risk factors that are genetic and not lifestyle related, a better understanding of their risks. This will also benefit policy making. For example, if the model identifies smoking as a strong predictor of a stroke, people in public health can prioritize initiatives that better control smoking and be more specfic in their messages regarding the long term effects.

## What future questions could this lead to?

One key area for future exploration is the generalizability of the model in real-word scenarios. This is specifically important in the event of insurance companies or governments deciding to use this technology in order to push their specific agendas. This may lead to unfair outcomes that discriminate or present biases based on the demographic of the original sample it was trained with. As refined and well-trained the model can be, it is crucial to consider the potential exceptions and variations that arise due to the oddities of human biology and environmental factors. Therefore this research can lead to future studies that focus on assessing the models ability across various populations. Other areas that can be explored are specific aspects of the predictors explored in this analysis. For example, one of our predictors is BMI, which is shown to be linked to all-cause mortality. However, BMI does not differentiate between fat mass and muscle mass (Cheung et al, 2015). Considering muscle mass is inversely associated with mortality (Lee & Giovannucci, 2018), researchers may want to see to what degree does fat mass associate with all-cause mortality, and in our specific case, stroke.

# References

Cheung, C., Lee, G. K., Au, P. C., Li, G. H., Chan, M., Li, H., Cheung, B. M., Wong, I. C., Lee, V. H., Mok, J., Yip, B. H., Cheng, K. K., & Wu, C. (2021). Systematic review and meta-analysis of lean mass and mortality: Rationale and study description. Osteoporosis and Sarcopenia, 7(Suppl 1), S3-S12. <https://doi.org/10.1016/j.afos.2021.01.001>

Lee, D. H., & Giovannucci, E. L. (2018). Body composition and mortality in the general population: A review of epidemiologic studies. Experimental biology and medicine (Maywood, N.J.), 243(17-18), 1275–1285. <https://doi.org/10.1177/1535370218818161>

Lev-Ari, S., Gepner, Y., & Goldbourt, U. (2021). Dissatisfaction with Married Life in Men Is Related to Increased Stroke and All-Cause Mortality. Journal of clinical medicine, 10(8), 1729. <https://doi.org/10.3390/jcm10081729>

Liu, X., Liu, C., Schenck, H., Yi, X., Wang, H., & Shi, X. (2017). The risk factors of 9-year follow-up on hypertension in middle-aged people in Tujia-Nationality settlement of China. Journal of human hypertension, 31(12), 838–842. <https://doi.org/10.1038/jhh.2017.58>

NCI Dictionary of Cancer terms. National Cancer Institute. (n.d.). <http://www.cancer.gov/publications/dictionaries/cancer-terms/def/stroke>

Shah, R. S., & Cole, J. W. (2010). Smoking and stroke: the more you smoke the more you stroke. Expert review of cardiovascular therapy, 8(7), 917–932. <https://doi.org/10.1586/erc.10.56>

Wang, S., Shen, B., Wu, M., Chen, C., & Wang, J. (2019). Effects of socioeconomic status on risk of ischemic stroke: A case-control study in the guangzhou population. BMC Public Health, 19(1), 648-648. <https://doi.org/10.1186/s12889-019-6998-4>
