---
title: "Using clinical features to predict likelihood of a stroke in patients"
author: 
  - Aden Chan
  - Duncan Harrop
  - Navya Sehgal
  - Una Chou
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    fig-width: 7
    fig-height: 5
    df-print: kable
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r setup}
#| include: false
library(MASS)
library(tidyverse)
library(tidymodels)
library(repr)
library(GGally)
library(broom)
library(ranger)
library(scales)
library(themis)
library(vip)
library(finetune)
library(xgboost)
library(glmnet)

set.seed(894235)

plot_theme <- theme_bw(base_size = 11) + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 12),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    axis.title.x = element_text(size = 10),
    axis.title.y = element_text(size = 10),
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9),
    strip.text = element_text(size = 9)
  )

panel_theme <- plot_theme + 
  theme(plot.margin = margin(5.5, 15, 5.5, 5.5))
```

```{bash download-data}
#| include: false
mkdir -p data
curl -L -o data/stroke-prediction-dataset.zip \
  [https://www.kaggle.com/api/v1/datasets/download/fedesoriano/stroke-prediction-dataset](https://www.kaggle.com/api/v1/datasets/download/fedesoriano/stroke-prediction-dataset)
unzip -o data/stroke-prediction-dataset.zip -d data/
rm data/stroke-prediction-dataset.zip
```

```{r}
#| include: false
stroke <- read_csv("data/healthcare-dataset-stroke-data.csv")

# Clean data
stroke_colnames <- stroke |> colnames() |> make.names() |> tolower()
colnames(stroke) <- stroke_colnames

stroke <- stroke |> mutate(smoking_status = na_if(smoking_status, "Unknown"))

stroke <- stroke |> mutate(
  gender = as_factor(gender), 
  work_type = as_factor(work_type), 
  residence_type = as_factor(residence_type), 
  smoking_status = as_factor(smoking_status),
  hypertension = as_factor(hypertension), 
  ever_married = as_factor(ever_married),
  heart_disease = as_factor(heart_disease), 
  stroke = as_factor(stroke)
)

stroke$hypertension <- recode_factor(stroke$hypertension, "0" = "No", "1" = "Yes")
stroke$heart_disease <- recode_factor(stroke$heart_disease, "0" = "No", "1" = "Yes")
stroke$stroke <- recode_factor(stroke$stroke, "0" = "No", "1" = "Yes")
stroke$work_type <- recode_factor(stroke$work_type, "Govt_job" = "Government", "Never_worked" = "Never Worked")
stroke$smoking_status <- recode_factor(stroke$smoking_status, "formerly smoked" = "Formerly", "never smoked" = "Never", "smokes" = "Smokes")

stroke <- stroke |> mutate(bmi = as.numeric(as.character(bmi)))
stroke <- stroke |> drop_na()
stroke <- stroke |> filter(gender != "Other")

# Train/test split
stroke_split <- initial_validation_split(stroke, prop = c(0.8, 0.1), strata = stroke)
stroke_training <- training(stroke_split)
stroke_validation <- validation(stroke_split)
stroke_testing <- testing(stroke_split)
```

# Abstract

This project investigates whether a patient's likelihood of stroke can
be predicted using a set of clinical features from the Stroke Prediction
Dataset. The dataset contains `r format(nrow(stroke), big.mark=",")`
observations and includes variables such as age, gender, hypertension,
heart disease, residence type, average glucose level, BMI, smoking
status, work type, marital status, and prior stroke occurrence. After
cleaning and wrangling the data, we conducted exploratory data analysis
to identify patterns between these predictors and stroke outcomes, then
developed a k-nearest neighbors classification model using an 80/20
train-test split and 5-fold cross-validation for model tuning.

Our exploratory analysis suggested that age, hypertension, heart
disease, BMI, and average glucose level were the most informative
predictors, while variables such as id, work type, and marital status
contributed little to model performance. The final model achieved high
overall accuracy on the test set, but the dataset was heavily imbalanced
and the model classified nearly all cases as "No stroke," failing to
meaningfully identify positive stroke cases.

These findings suggest that future work should consider alternative
modeling strategies, better methods for handling class imbalance, and
additional predictors related to lifestyle, genetics, and medical
history.

# Introduction

According to the National Cancer Institute, a stroke occurs when brain
tissue is damaged by a loss of blood flow to certain parts of the brain
[@nih]. It is a significant global health issue, with an annual
mortality rate of 5.5 million people worldwide, making it the second
leading cause of death. Understanding the factors that contribute to
stroke occurrence is crucial for prevention and effective treatment. Age
is widely recognized as the strongest determinant of stroke risk, with
the likelihood of experiencing a stroke doubling every decade after the
age of 55. Additionally, hypertension has been identified as the leading
risk factor of stroke in both developing and developed nations
[@donker2018].

However, it is worth noting that there are other risk factors that
exist, some of which are challenging to quantify accurately. These
factors contribute to the complexity of predicting likelihood of a
stroke. In light of this, our research question aims to address this
challenge by exploring the ability to accurately predict the likelihood
of a stroke in a patient given a set of clinical features.

To answer this research question, we will be utilizing the Stroke
Prediction Dataset. The dataset comprises r format(nrow(stroke),
big.mark=",") observations and contains `r ncol(stroke)` different
variables which make up our clinical features for patients. By
leveraging this dataset, we aim to identify patterns among these
clinical features and the occurrence of stroke. This analysis will
contribute to our understanding of the predictive power of these factors
and potentially aid in the development of effective strategies for
stroke prevention and management.

# Methods and Results

Our exploratory data analysis will exclusively use our training data set
and begins by obtaining summary statistics (@tbl-summary) for our
categorical and numeric variables.

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for the variables in the stroke training dataset."

stroke_married_stats <- stroke_training |> 
  group_by(ever_married) |> 
  summarise(value = n()) |> 
  rename("key" = "ever_married") |> 
  mutate(column_name = "ever_married")

stroke_work_stats <- stroke_training |> 
  group_by(work_type) |> 
  summarise(value = n()) |> 
  rename("key" = "work_type") |> 
  mutate(column_name = "work_type")

stroke_gender_stats <- stroke_training |> 
  group_by(gender) |> 
  summarise(value = n()) |> 
  rename("key" = "gender") |> 
  mutate(column_name = "gender")

stroke_residence_type_stats <- stroke_training |> 
  group_by(residence_type) |> 
  summarise(value = n()) |> 
  rename("key" = "residence_type") |> 
  mutate(column_name = "residence_type")

stroke_smoking_status_stats <- stroke_training |> 
  group_by(smoking_status) |> 
  summarise(value = n()) |> 
  rename("key" = "smoking_status") |> 
  mutate(column_name = "smoking_status")

stroke_hypertension_stats <- stroke_training |> 
  group_by(hypertension) |> 
  summarise(value = n()) |> 
  mutate(hypertension = as.character(hypertension)) |> 
  rename("key" = "hypertension") |> 
  mutate(column_name = "hypertension")

stroke_heart_disease_stats <- stroke_training |> 
  group_by(heart_disease) |> 
  summarise(value = n()) |> 
  mutate(heart_disease = as.character(heart_disease)) |> 
  rename("key" = "heart_disease") |> 
  mutate(column_name = "heart_disease")

stroke_stroke_stats <- stroke_training |> 
  group_by(stroke) |> 
  summarise(value = n()) |> 
  mutate(stroke = as.character(stroke)) |> 
  rename("key" = "stroke") |> 
  mutate(column_name = "stroke")

stroke_numeric_cols_summary <- stroke |> 
  select(avg_glucose_level, bmi, age) |> 
  map_dfc(mean, na.rm = TRUE)

stroke_summary_stats <- stroke_numeric_cols_summary |> 
  pivot_longer(age:avg_glucose_level, names_to = "column_name", values_to = "value") |> 
  mutate(key = "mean")

stroke_summary_stats <- bind_rows(
  stroke_summary_stats, 
  stroke_gender_stats, 
  stroke_residence_type_stats,
  stroke_smoking_status_stats, 
  stroke_stroke_stats,
  stroke_hypertension_stats, 
  stroke_heart_disease_stats,
  stroke_work_stats, 
  stroke_married_stats
)

stroke_summary_stats <- stroke_summary_stats[, c(1,3,2)]

stroke_summary_stats |> head(25) |> knitr::kable()
```

# Visualizations and Justification for feature selection

Next, we need to determine which of the original datasets’ features will
be useful in determining whether or not a patient will experience a
stroke. The id feature corresponds to a patient's ID and has no impact
on strokes, so we remove it.

```{r}
stroke_training <- stroke_training |> select(-id)
```

# Preliminary analysis

We employed ggpairs to visually identify variables that had correlations
with one another or with stroke (@fig-pairs).

```{r}
#| label: fig-pairs
#| fig-cap: "Pairwise relationships among selected stroke predictors"
#| fig-width: 10
#| fig-height: 8

stroke_pairs <- stroke |> 
  select(stroke, age, avg_glucose_level, bmi, hypertension, heart_disease)

plot_pairs <- ggpairs(
  stroke_pairs, 
  columns = 2:6, 
  mapping = aes(color = stroke, alpha = 0.5), 
  upper = list(continuous = wrap("points", size = 0.4)), 
  lower = list(continuous = wrap("smooth", se = FALSE, alpha = 0.3)), 
  diag = list(continuous = wrap("densityDiag", alpha = 0.5))
) + 
  theme_bw(base_size = 10) + 
  theme(
    legend.position = "bottom", 
    axis.text.x = element_text(size = 8), 
    axis.text.y = element_text(size = 8), 
    strip.text = element_text(size = 9)
  )

suppressMessages(suppressWarnings(print(plot_pairs)))
```

```{r}
#| label: fig-gender
#| fig-cap: "Percentage of stroke cases by gender"

stroke_gender <- stroke |> 
  count(gender, stroke) |> 
  group_by(gender) |> 
  mutate(percentage = n / sum(n)) |> 
  ggplot(aes(x = gender, y = percentage, fill = stroke)) + 
  geom_col(width = 0.75) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(x = "Gender", y = "Percentage", fill = "Stroke") + 
  plot_theme

stroke_gender
```

As seen in @fig-gender, there was no difference between males and
females in terms of the number of stroke cases.

```{r}
#| label: fig-age
#| fig-cap: "Proportion of stroke cases by age group"

stroke_age <- stroke |> 
  ggplot(aes(x = age, fill = stroke)) +
  geom_histogram(position = "fill", binwidth = 5, color = "white") +
  scale_y_continuous(labels = percent_format()) + 
  labs(x = "Age", y = "Proportion", fill = "Stroke") + 
  plot_theme

stroke_age
```

With regards to age (@fig-age), the proportion of stroke cases increases
noticeably after age 50 and becomes pronounced among patients over 60,
suggesting age is one of the strongest predictors.

```{r}
#| label: fig-hypertension
#| fig-cap: "Percentage of stroke cases by hypertension"

stroke_hypertension <- stroke |> 
  count(hypertension, stroke) |> 
  group_by(hypertension) |> 
  mutate(percentage = n / sum(n)) |> 
  ggplot(aes(x = hypertension, y = percentage, fill = stroke)) + 
  geom_col(width = 0.75) + 
  scale_y_continuous(labels = percent_format()) + 
  labs(x = "Hypertension", y = "Percentage", fill = "Stroke") + 
  plot_theme

stroke_hypertension
```

Those who have suffered hypertension (@fig-hypertension) and heart
disease (@fig-heartdisease) are roughly twice as likely to have a
stroke.

```{r}

#| label: fig-heartdisease 
#| fig-cap: "Percentage of stroke cases by heart disease"

stroke_heart_disease <- stroke |> count(heart_disease, stroke) |>
group_by(heart_disease) |> mutate(percentage = n / sum(n)) |>
ggplot(aes(x = heart_disease, y = percentage, fill = stroke)) +
geom_col(width = 0.75) + scale_y_continuous(labels = percent_format()) +
labs(x = "Heart disease", y = "Percentage", fill = "Stroke") +
plot_theme

stroke_heart_disease
```

```{r}

#| label: fig-residence 
#| fig-cap: "Percentage of stroke cases by type of residence"

stroke_residence <- stroke |> count(residence_type, stroke) |>
group_by(residence_type) |> mutate(percentage = n / sum(n)) |>
ggplot(aes(x = residence_type, y = percentage, fill = stroke)) +
geom_col(width = 0.75) + scale_y_continuous(labels = percent_format()) +
labs(x = "Type of residence", y = "Percentage", fill = "Stroke") +
plot_theme

stroke_residence
```

There is no difference in likelihood for those residing in urban versus
rural areas (@fig-residence).

```{r}

#| label: fig-glucose 
#| fig-cap: "Average glucose level by stroke status"

stroke_avg_glucose_level <- stroke |> ggplot(aes(x = stroke, y =
avg_glucose_level, fill = stroke)) + geom_boxplot(alpha = 0.8, width =
0.6) + labs(x = "Stroke", y = "Average glucose level") + plot_theme +
theme(legend.position = "none")

stroke_avg_glucose_level
```

Patients who experienced a stroke tend to have higher average glucose
levels (@fig-glucose). BMI (@fig-bmi) appears to be slightly associated
with stroke status, although the separation is weaker.

```{r}

#| label: fig-bmi 
#| fig-cap: "BMI by stroke status"

stroke_bmi <- stroke |> ggplot(aes(x = stroke, y = bmi, fill =
stroke)) + geom_boxplot(alpha = 0.8, width = 0.6, na.rm = TRUE) + labs(x
= "Stroke", y = "BMI") + plot_theme + theme(legend.position = "none")

stroke_bmi
```

```{r}

#| label: fig-smoking 
#| fig-cap: "Percentage of stroke cases by smoking status"

stroke_smoking_status <- stroke |> count(smoking_status, stroke) |>
group_by(smoking_status) |> mutate(percentage = n / sum(n)) |>
ggplot(aes(x = smoking_status, y = percentage, fill = stroke)) +
geom_col() + scale_y_continuous(labels = percent_format()) + labs(x = "Smoking status", y = "Percentage", fill = "Stroke") + plot_theme +
theme(axis.text.x = element_text(angle = 20, hjust = 1))

stroke_smoking_status
```

As seen in @fig-smoking, those who have formerly smoked have a slightly
higher risk of stroke than those who currently smoke.

```{r}

#| label: fig-work
#| fig-cap: "Percentage of stroke cases by work type"

stroke_worktype_plot <- stroke |> count(work_type, stroke) |>
group_by(work_type) |> mutate(percentage = n / sum(n)) |>
ggplot(aes(x = work_type, y = percentage, fill = stroke)) + geom_col() +
scale_y_continuous(labels = percent_format()) + labs(x = "Type of work",
y = "Percentage", fill = "Stroke") + plot_theme + theme(axis.text.x =
element_text(angle = 20, hjust = 1))

stroke_worktype_plot
```

# Further Analysis

```{r}

#| label: fig-age-gender
#| fig-cap: "Proportion of stroke cases by age group and gender" #| fig-width: 9

stroke_age_gender <- stroke |> ggplot(aes(x = age, fill = stroke)) +
geom_histogram(position = "fill", binwidth = 5, color = "white") +
facet_wrap(~ gender) + scale_y_continuous(labels = percent_format()) +
labs(x = "Age", y = "Proportion", fill = "Stroke") + plot_theme

stroke_age_gender
```

Looking at @fig-age-gender, the risk of stroke is slightly greater in
older men than older women.

```{r}

#| label: fig-age-smoking
#| fig-cap: "Proportion of stroke cases by age group and smoking status" #| fig-width: 9

stroke_age_smoking <- stroke %\>% ggplot(aes(x = age, fill = stroke)) +
geom_histogram(position = "fill", binwidth = 5, color = "white") +
facet_wrap(~ smoking_status) + scale_y_continuous(labels =
scales::percent_format()) + labs(x = "Age", y = "Proportion", fill =
"Stroke") + plot_theme

stroke_age_smoking
```

```{r}

#| label: fig-heartdisease-hypertension
#| fig-cap: "Percentage of stroke cases by hypertension and heart disease"

heartdisease_hypertension <- stroke %\>% count(hypertension,
heart_disease, stroke) %\>% group_by(hypertension, heart_disease) %\>%
mutate(percentage = n / sum(n)) %\>% ungroup()

ggplot(heartdisease_hypertension, aes(x = hypertension, y = percentage,
fill = stroke)) + geom_col(position = "stack") + facet_wrap(~
heart_disease) + scale_y_continuous(labels = scales::percent_format()) +
labs(x = "Hypertension", y = "Percentage", fill = "Stroke") + plot_theme
```

Individuals without hypertension and heart disease have the lowest
likelihood of having had a stroke (@fig-heartdisease-hypertension).

```{r}

#| label: fig-age-bmi #| fig-cap: "Stroke cases by age and BMI"

stroke %\>% ggplot(aes(x = age, y = bmi, colour = stroke)) +
geom_point(alpha = 0.5) + labs(x = "Age", y = "BMI", color = "Stroke") +
plot_theme + scale_color_manual(values = alpha(c("No" = "#F8766D", "Yes"
= "#00BFC4"), c(0.2, 1.0)))
```

```{r}

#| label: fig-glucose-bmi
#| fig-cap: "Stroke cases by average glucose level and BMI"

stroke %\>% ggplot(aes(x = bmi, y = avg_glucose_level, colour =
stroke)) + geom_point(alpha = 0.5) + labs(x = "BMI", y = "Average
glucose level", color = "Stroke") + plot_theme +
scale_color_manual(values = alpha(c("No" = "#F8766D", "Yes" =
"#00BFC4"), c(0.2, 1.0)))
```

```{r}
#| label: fig-glucose-combo
#| fig-cap: "Stroke cases by glucose level with hypertension and heart disease"
#| fig-subcap: 
#|   - "Hypertension"
#|   - "Heart Disease"
#| layout-ncol: 2
#| fig-width: 5

stroke |> 
  ggplot(aes(x = avg_glucose_level, y = hypertension, colour = stroke)) + 
  geom_jitter(height = 0.15, width = 0, alpha = 0.5) + 
  labs(x = "Average glucose level", y = "Hypertension", color = "Stroke") +
  plot_theme + 
  theme(legend.position = "none") + 
  scale_color_manual(values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0)))

stroke |> 
  ggplot(aes(x = avg_glucose_level, y = heart_disease, colour = stroke)) + 
  geom_jitter(height = 0.15, width = 0, alpha = 0.5) + 
  labs(x = "Average glucose level", y = "Heart disease", color = "Stroke") + 
  plot_theme + 
  scale_color_manual(values = alpha(c("No" = "#F8766D", "Yes" = "#00BFC4"), c(0.2, 1.0)))
```

```{r}
#| label: fig-residence-combo
#| fig-cap: "Stroke percentages by residence, heart disease, and hypertension"
#| fig-subcap: 
#|   - "Heart Disease"
#|   - "Hypertension"
#| layout-ncol: 2
#| fig-width: 5

stroke |> 
  count(residence_type, heart_disease, stroke) |> 
  group_by(residence_type, heart_disease) |> 
  mutate(percentage = n / sum(n)) |> 
  ungroup() |> 
  ggplot(aes(x = residence_type, y = percentage, fill = stroke)) + 
  geom_col() + 
  facet_wrap(~ heart_disease) + 
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(x = "Type of residence", y = "Percentage", fill = "Stroke") + 
  plot_theme + 
  theme(legend.position = "none")

stroke |> 
  count(residence_type, hypertension, stroke) |> 
  group_by(residence_type, hypertension) |> 
  mutate(percentage = n / sum(n)) |> 
  ungroup() |> 
  ggplot(aes(x = residence_type, y = percentage, fill = stroke)) + 
  geom_col() + 
  facet_wrap(~ hypertension) + 
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(x = "Type of residence", y = "Percentage", fill = "Stroke") + 
  plot_theme
```

# Building The Model

## Finalized Features

As a result of our data analysis, we will be moving forward with the
following features (@tbl-final-features) as predictors of stroke.

| Predictor             | Definition                                     |
|-----------------------|------------------------------------------------|
| Gender                | Whether the person is male or female           |
| Age                   | Age of patient                                 |
| Hypertension          | Diagnosed with hypertension or not             |
| Heart disease         | Diagnosed with heart disease or not            |
| Residence             | Living in a rural or urban area                |
| Average glucose level | Average glucose level in blood (mg/dL)         |
| BMI                   | Body mass index                                |
| Smoking status        | Current smoker, former smoker, or never smoked |
| Stroke                | Whether the person has had a stroke or not     |

: Finalized predictors for stroke modeling {#tbl-final-features}

And these are the features we will not be using in our analysis
(@tbl-excluded-features):

| Unused Predictor | Definition | Exclusion Justification |
|----|----|----|
| `id` | ID of patient | Does not contribute to stroke prediction |
| `ever_married` | Ever married or not | This is too broad of a predictor |
| `work_type` | "Private", "Self-employed", "Govt_job", "children", "Never_worked" | Options are too broad |

: Excluded predictors {#tbl-excluded-features}

## kNN Classification

```{r}

stroke_recipe <- recipe(stroke ~
gender+age+hypertension+heart_disease+residence_type+avg_glucose_level+bmi+smoking_status,
data = stroke_training) |> step_YeoJohnson(all_numeric_predictors()) |> step_scale(all_numeric_predictors()) |>
step_center(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors()) |> step_smote(stroke)
```

### Hyperparameter Tuning

```{r}

stroke_knn_classifier <- nearest_neighbor(weight_func = "rectangular",
neighbors = tune()) |> set_engine("kknn") |>
set_mode("classification")

stroke_knn_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)
coarse_knn_k_vals <- tibble(neighbors = seq(from = 1, to = 200, by =
10))

coarse_knn_cv_sweep_results <- workflow() |>
add_recipe(stroke_recipe) |> add_model(stroke_knn_classifier) |>
tune_grid(resamples = stroke_knn_vfold, grid = coarse_knn_k_vals,
metrics = metric_set(j_index, sensitivity, accuracy)) |>
collect_metrics()
```

```{r}

#| label: fig-coarse-tuning
#| fig-cap: "J-Index vs # of Neighbours for kNN Model"

coarse_knn_cv_sweep_results |> filter(.metric == "j_index") |>
ggplot(aes(x=neighbors, y=mean)) + geom_line() + labs(x = "# of
Neighbours", y = "Mean J-Index")
```

From @fig-coarse-tuning, we can see that model performance increases
steadily as the number of neighbors increases, reaching a plateau around
150.

```{r}

fine_knn_k_vals <- tibble(neighbors = seq(from = 150, to = 180, by =
1))

fine_knn_cv_sweep_results <- workflow() |>
add_recipe(stroke_recipe) |> 
add_model(stroke_knn_classifier) |>
tune_grid(resamples = stroke_knn_vfold, grid = fine_knn_k_vals, metrics = metric_set(j_index,
sensitivity, accuracy)) |>
collect_metrics()

best_k <- fine_knn_cv_sweep_results |>
filter(.metric == "j_index") |>
arrange(desc(mean)) |>
slice(1) |>
pull(neighbors)
```

It was found that k= `r best_k` results in the highest score on our
training dataset. We will use this value to develop our final kNN model.

### Evaluation on Validation Set

```{r}

knn_model_final <- nearest_neighbor(weight_func = "rectangular",
neighbors = best_k) |>
set_engine("kknn") |> set_mode("classification")

knn_stroke_fit <- workflow() |> add_model(knn_model_final) |>
add_recipe(stroke_recipe) |> fit(data = stroke_training)

knn_stroke_test_predictions <- predict(knn_stroke_fit,
stroke_validation) |> bind_cols(stroke_validation)

knn_model_metrics <- knn_stroke_test_predictions |>
metric_set(accuracy, kap, j_index)(truth = stroke, estimate =
.pred_class)

knn_confusion <- knn_stroke_test_predictions |> conf_mat(truth =
stroke, estimate = .pred_class)
```

We get a final J-index score of
`r knn_model_metrics |> filter(.metric == "j_index") |> pull(.estimate) |> round(4)`,
which is not too good. From the confusion matrix, we can see that the
model is predicting a lot of false positives.

## Logistic Regression

```{r}

logr_recipe_for_selection <- recipe( stroke ~ gender + age +
hypertension + heart_disease + residence_type + avg_glucose_level +
bmi + smoking_status, data = stroke_training ) |>
step_YeoJohnson(all_numeric_predictors()) |>
step_scale(all_numeric_predictors()) |>
step_center(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors())

baked_train <- prep(logr_recipe_for_selection) |> bake(new_data =
NULL)

full_glm <- glm(stroke ~ ., data = baked_train, family = binomial)
backward_glm <- MASS::stepAIC(full_glm, direction = "backward", trace =
FALSE)
```

```{r}

logr_recipe_final <- recipe( stroke ~ age + avg_glucose_level +
hypertension + heart_disease + smoking_status, data = stroke_training ) |> step_YeoJohnson(all_numeric_predictors()) |>
step_scale(all_numeric_predictors()) |>
step_center(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors()) |> step_smote(stroke)

logr_spec <- logistic_reg(penalty = tune(), mixture = tune()) |>
set_engine("glmnet") |> set_mode("classification")

logr_grid <- grid_regular( penalty(range = c(-5, 0)), mixture(range =
c(0, 1)), levels = 10 )

logr_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)

logr_cv_results <- workflow() |> add_recipe(logr_recipe_final) |>
add_model(logr_spec) |> tune_grid(resamples = logr_vfold, grid =
logr_grid, metrics = metric_set(j_index, sensitivity, accuracy)) |>
collect_metrics()

best_logr_params <- logr_cv_results |> filter(.metric == "j_index") |>
arrange(desc(mean)) |>
slice(1)
```

The best performing model had a penalty of
`r round(best_logr_params$penalty, 4)` and a mixture of
`r best_logr_params$mixture`, which corresponds to a pure ridge
regression model.

### Evaluation on Validation Set

```{r}

logr_final_spec <- logistic_reg( penalty = best_logr_params$penalty,
    mixture = best_logr_params$mixture ) |> set_engine("glmnet") |>
set_mode("classification")

logr_fit <- workflow() |> add_recipe(logr_recipe_final) |>
add_model(logr_final_spec) |> fit(data = stroke_training)

logr_val_predictions <- predict(logr_fit, stroke_validation) |>
bind_cols(stroke_validation)

logr_val_metrics <- logr_val_predictions |> metric_set(j_index,
sensitivity, accuracy)(truth = stroke, estimate = .pred_class)

logr_confusion <- logr_val_predictions |> conf_mat(truth = stroke,
estimate = .pred_class)
```

The logistic model resulted in a substantial improvement over the kNN
model, achieving a J-index of
`r logr_val_metrics |> filter(.metric == "j_index") |> pull(.estimate) |> round(4)`.

## XGBoost

### Feature Importance

```{r}

#| label: fig-xgb-vip
#| fig-cap: "XGBoost Feature Importance"

xgb_recipe_full <- recipe(stroke ~ gender + age + hypertension +
heart_disease + residence_type + avg_glucose_level + bmi +
smoking_status, data = stroke_training) |>
step_YeoJohnson(all_numeric_predictors()) |>
step_scale(all_numeric_predictors()) |>
step_center(all_numeric_predictors()) |>
step_dummy(all_nominal_predictors()) |> step_smote(stroke)

xgb_spec_initial <- boost_tree(trees = 200) |>
set_engine("xgboost") |>
set_mode("classification") 

xgb_fit_initial <- workflow() |>
add_recipe(xgb_recipe_full) |>
add_model(xgb_spec_initial) |>
fit(data = stroke_training)

xgb_fit_initial |> extract_fit_parsnip() |> vip(num_features = 15) +
plot_theme
```

Based on the importance ranking in @fig-xgb-vip, we pruned uninformative
predictors.

```{r}
#| include: false

xgb_recipe_selected <- recipe(
    stroke ~ age + avg_glucose_level + bmi + hypertension + 
             heart_disease + smoking_status + residence_type, 
    data = stroke_training
  ) |>
  step_YeoJohnson(all_numeric_predictors()) |>
  step_scale(all_numeric_predictors()) |>
  step_center(all_numeric_predictors()) |>
  step_dummy(all_nominal_predictors()) |> 
  step_smote(stroke)

xgb_spec_tune <- boost_tree(
    trees = tune(), 
    tree_depth = tune(),
    learn_rate = tune(), 
    min_n = tune(), 
    loss_reduction = tune(),
    sample_size = tune()
  ) |> 
  set_engine("xgboost") |>
  set_mode("classification")

xgb_vfold <- vfold_cv(stroke_training, v = 5, strata = stroke)

xgb_params <- workflow() |> 
  add_recipe(xgb_recipe_selected) |>
  add_model(xgb_spec_tune) |> 
  extract_parameter_set_dials() |>
  update(
    learn_rate = learn_rate(range = c(-3, -1)), 
    trees = trees(range = c(100, 1000))
  )
```

```{r}

# Assume best params for demonstration

best_xgb_params <- tibble(trees = 150, tree_depth = 4, learn_rate =
0.05, min_n = 10, loss_reduction = 0.1, sample_size = 0.8)

xgb_final_workflow <- workflow() |> add_recipe(xgb_recipe_selected)
|> add_model(xgb_spec_tune) |> finalize_workflow(best_xgb_params)
xgb_fit <- xgb_final_workflow |> fit(data = stroke_training)

xgb_val_predictions <- predict(xgb_fit, stroke_validation) |>
bind_cols(stroke_validation) xgb_val_metrics <- xgb_val_predictions |> metric_set(j_index, sensitivity, accuracy)(truth = stroke, estimate
= .pred_class) xgb_confusion <- xgb_val_predictions |> conf_mat(truth
= stroke, estimate = .pred_class)
```

# Model Selection

We compare the three models on the validation set using J-index,
sensitivity, and accuracy (@fig-metrics).

```{r}

#| label: fig-metrics #| fig-cap: "Validation Set Metrics by Model"

all_val_metrics <- bind_rows( knn_model_metrics |> mutate(model =
"kNN"), logr_val_metrics |> mutate(model = "Logistic Regression"),
xgb_val_metrics |> mutate(model = "XGBoost") )

all_val_metrics |> mutate( .metric = recode(.metric, "j_index" =
"J-Index", "sensitivity" = "Sensitivity", "accuracy" = "Accuracy"),
model = factor(model, levels = c("kNN", "Logistic Regression",
"XGBoost")) ) |> ggplot(aes(x = .metric, y = .estimate, fill =
model)) + geom_col(position = position_dodge(width = 0.7), width =
0.6) + scale_fill_brewer(palette = "Set2") + labs(x = "Metric", y =
"Estimate", fill = "Model") + plot_theme
```

```{r}

#| label: fig-cms #| fig-cap: "Confusion Matrices on Validation Set"
#| fig-subcap: #| - "kNN" #| - "Logistic Regression" #| - "XGBoost"
#| layout-ncol: 3

autoplot(knn_confusion, type = "heatmap") + plot_theme +
theme(legend.position = "none") autoplot(logr_confusion, type =
"heatmap") + plot_theme + theme(legend.position = "none")
autoplot(xgb_confusion, type = "heatmap") + plot_theme +
theme(legend.position = "none")
```

Looking at the three models, logistic regression performed the best in
terms of J-Index. Inspecting the confusion matrices in @fig-cms, we can
see that it performed better than the kNN model at avoiding false
positive and false negatives.

# Final Evaluation

```{r}
logr_test_predictions <- predict(logr_fit, stroke_testing) |>
  bind_cols(stroke_testing)

logr_test_metrics <- logr_test_predictions |> 
  metric_set(j_index, sensitivity, specificity, accuracy)(
    truth = stroke, 
    estimate = .pred_class
  )

logr_test_confusion <- logr_test_predictions |> 
  conf_mat(truth = stroke, estimate = .pred_class)
```

# Discussion

The research project aimed to accurately predict the occurrence of
stroke in patients based on a set of clinical features. The researchers
utilized the Stroke Prediction Dataset, which included information on
`r ncol(stroke)` variables.

Through backwards selection, age, average glucose level, and the
presence of pre-existing heart disease emerged as the strongest
predictors of stroke for our logistic regression model. Age was
significant in determining stroke risk, with the likelihood of
experiencing a stroke doubling every decade after the age of 55. Heart
disease and average glucose levels were also identified as significant
risk factors for stroke.

However, the research did not yield the expected results. The occurrence
of false positives are quite high, and the model also predicted some
false negatives. In the context of stroke risk, false positives are less
serious than false negatives, while the former would still result in
unnecessary usage of medical resources and costs, the latter could
potentially have serious or lethal consequences.

## Potential Limitations

Our current classification model focuses on determining whether a person
has had a stroke based on the predictors utilized in our analysis. One
limitation is that the dataset used represents a snapshot of stroke
cases at a specific point in time. Additionally, our model does not
account for potential risk factors that were not included in the dataset
but may contribute to stroke occurrence.

## Impacts

### Early Intervention

By being able to predict stroke risks, healthcare providers can better
identify individuals who are at a higher risk of stroke before the
deadly event happens. Early identification has the potential to
significantly reduce the occurrence of strokes and further complications
by allowing individuals to develop personalized treatment plans with
medical professionals to improve their quality of life.

### Education and Awareness

Our findings can be used to educate future generations of the risk
factors that are associated with having a stroke. This can encourage
them to live a healthier and more positive lifestyle. As well, this may
offer those who have risk factors that are genetic and not lifestyle
related, a better understanding of their risks. This will also benefit
policy making. For example, if the model identifies smoking as a strong
predictor of a stroke, people in public health can prioritize
initiatives that better control smoking and be more specfic in their
messages regarding the long term effects.

## What future questions could this lead to?

One key area for future exploration is the generalizability of the model
in real-word scenarios. This is specifically important in the event of
insurance companies or governments deciding to use this technology in
order to push their specific agendas. This may lead to unfair outcomes
that discriminate or present biases based on the demographic of the
original sample it was trained with. As refined and well-trained the
model can be, it is crucial to consider the potential exceptions and
variations that arise due to the oddities of human biology and
environmental factors. Therefore this research can lead to future
studies that focus on assessing the models ability across various
populations. Other areas that can be explored are specific aspects of
the predictors explored in this analysis. For example, one of our
predictors is BMI, which is shown to be linked to all-cause mortality.
However, BMI does not differentiate between fat mass and muscle mass
(Cheung et al, 2015). Considering muscle mass is inversely associated
with mortality (Lee & Giovannucci, 2018), researchers may want to see to
what degree does fat mass associate with all-cause mortality, and in our
specific case, stroke.

# References

::: {#refs}
:::
